### 2025-06-06 basic-statistics.qmd
20250606 - update formulas

The mean of $X$ (written as: $\bar{X}$) is calculated as in @eq-IQR:

$$
\bar{X} = \frac{\sum X_i}{n} = \frac{X_1 + X_2 + ... + X_n}{n}
$$ {#eq-IQR}

@@

The (sample) variance of $X$ (written as: $s^2$) is calculated as in @eq-variance:

$$
s^2 = \frac{\sum (X_i - \bar{X})^2}{n-1}
$$ {#eq-variance}

where $X_i$ is each data point, $\bar{X}$ is the mean of $X$, and $n$ is the number of data points.

The (sample) standard deviation of $X$ (written as: $s$) is calculated as in @eq-sd:

$$
s = \sqrt{\frac{\sum (X_i - \bar{X})^2}{n-1}}
$$ {#eq-sd}

The range is calculated of $X$ is calculated as in @eq-range:

@@

$$
\text{MAD} = \text{median}(|X_i - \tilde{X}|)
$$ {#eq-medianAbsoluteDeviation}

where $\tilde{X}$ is the median of `X`.
Compared to the standard deviation, the median absolute deviation is more robust to outliers.

@@

$$
z = \frac{X - \bar{X}}{\sigma}
$$ {#eq-zScore}

where $X$ is the observed score, $\bar{X}$ is the mean observed score, and $\sigma$ is the standard deviation of the observed scores.

@@

$$
\begin{aligned}
  R^2 &= 1 - \frac{\sum (Y_i - \hat{Y}_i)^2}{\sum (Y_i - \bar{Y})^2} \\
      &= 1 - \frac{SS_{\text{residual}}}{SS_{\text{total}}} \\
      &= 1 - \frac{\text{sum of squared residuals}}{\text{total sum of squares}} \\
      &= \frac{f^2}{1 + f^2} \\
      &= \eta^2 \\
      &= \frac{\text{variance explained in }Y}{\text{total variance in }Y}
\end{aligned}
$$ {#eq-rSquared}

where $Y_i$ is the observed value of the [outcome variable](#sec-correlationalStudy) for the $i$th observation, $\hat{Y}_i$ is the model predicted value for the $i$th observation, $\bar{Y}$ is the mean of the observed values of the [outcome variable](#sec-correlationalStudy).
The total sum of squares is an index of the total variation in the [outcome variable](#sec-correlationalStudy).

@@

where $X$ is the [predictor variable](#sec-correlationalStudy) and $Y$ is the [outcome variable](#sec-correlationalStudy).

>>>>

The mean of $x$ (written as: $\bar{x}$) is calculated as in @eq-IQR:

$$
\bar{x} = \frac{\sum x_i}{n} = \frac{x_1 + x_2 + ... + x_n}{n}
$$ {#eq-IQR}

@@

The (sample) variance of $x$ (written as: $s^2$) is calculated as in @eq-variance:

$$
s^2 = \frac{\sum (x_i - \bar{x})^2}{n-1}
$$ {#eq-variance}

where $x_i$ is each data point, $\bar{x}$ is the mean of $x$, and $n$ is the number of data points.

The (sample) standard deviation of $x$ (written as: $s$) is calculated as in @eq-sd:

$$
s = \sqrt{\frac{\sum (x_i - \bar{x})^2}{n-1}}
$$ {#eq-sd}

The range is calculated of $x$ is calculated as in @eq-range:

@@

$$
\text{MAD} = \text{median}(|x_i - \tilde{x}|)
$$ {#eq-medianAbsoluteDeviation}

where $\tilde{x}$ is the median of $x$.
Compared to the standard deviation, the median absolute deviation is more robust to outliers.

@@

$$
z = \frac{x - \bar{x}}{\sigma}
$$ {#eq-zScore}

where $x$ is the observed score, $\bar{x}$ is the mean observed score, and $\sigma$ is the standard deviation of the observed scores.

@@

$$
\begin{aligned}
  R^2 &= 1 - \frac{\sum (y_i - \hat{y}_i)^2}{\sum (y_i - \bar{y})^2} \\
      &= 1 - \frac{SS_{\text{residual}}}{SS_{\text{total}}} \\
      &= 1 - \frac{\text{sum of squared residuals}}{\text{total sum of squares}} \\
      &= \frac{f^2}{1 + f^2} \\
      &= \eta^2 \\
      &= \frac{\text{variance explained in }y}{\text{total variance in }y}
\end{aligned}
$$ {#eq-rSquared}

where $y$ is the [outcome variable](#sec-correlationalStudy), $y_i$ is the observed value of the [outcome variable](#sec-correlationalStudy) for the $i$th observation, $\hat{y}_i$ is the model predicted value for the $i$th observation, $\bar{y}$ is the mean of the observed values of the [outcome variable](#sec-correlationalStudy).
The total sum of squares is an index of the total variation in the [outcome variable](#sec-correlationalStudy).

@@

where $x$ is the [predictor variable](#sec-correlationalStudy), $y$ is the [outcome variable](#sec-correlationalStudy), and $\text{Cov}(x, y)$ is the covariance between $x$ and $y$.
